Fusing Filters with Integer Linear Programming

### 2014-05-12 2014betterfusionforfilters/Main.tex
Rewrote the intro

\begin{abstract}
Minimising memory traffic is key to compiling functional, collection oriented array programs into efficient code. To achieve this, it is not sufficient to just fuse subsequent array operations into a single, more complex computation, we also have to also \emph{cluster} separate traversals of the same array into one single loop. Previous work demonstrated how Integer Linear Programming (ILP) can be used to cluster the operators in a general data-flow graph into subgraphs, which can be individually fused. However, these approaches have one serious restrictions: they can only handle operations which preserve the size of the array. The technique we present in this paper extends the ILP approach with support for size-changing operations, using an external ILP server to find good clusterings.

>>>>

\begin{abstract}
Minimising memory traffic is key to compiling functional, collection oriented array programs into efficient code. To achieve this, it is not sufficient to just fuse subsequent array operations into a single, more complex computation, we also have to also \emph{cluster} separate traversals of the same array into one single loop. Previous work demonstrated how Integer Linear Programming (ILP) can be used to cluster the operators in a general data-flow graph into subgraphs, which can be individually fused. However, these approaches have one serious restrictions: they can only handle operations which preserve the size of the array, thereby missing out on some optimisation opportunities. The technique we present in this paper addresses this shortcoming by extending the ILP approach with support for size-changing operations, using an external ILP server to find good clusterings.

### 2014-05-12 2014betterfusionforfilters/section/01-Introduction.tex
Rewrote the intro

To demonstrate the effect of clustering and fusion, consider the following code snippet:
\begin{code}
 normalize2 :: Array Int -> Array Int
 normalize2 xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (> 0)    xs
        sum2 = fold   (+)  0   gts
        ys1  = map    (/ sum1) xs
        ys2  = map    (/ sum2) xs
    in (ys1, ys2)
\end{code}
The function @normalize@ computes two sums, one of all the elements of @xs@, the other of all the ones greater than zero. Since we need to fully evalute the sums to proceed with the maps, it is clear that we need at least two separate loops. These folds are examples of fusion-preventing dependencies, as fold cannot be fused with subsequent operations. Figure~\ref{f:normalize2-clusterings} shows the data-flow graph of @normalize@. In the leftmost diagram we can see the effect of applying stream fusion to the program: we end up with four loops (denoted by the dotted lines): only the filter operation is combined with the subsequent fold. The best existing ILP approach results in the right-most graph: it combines the @sum1@ fold and @sum2@ filter in one loop, but requires an extra loop for the fold operation which consumes the filter output, since it cannot fuse filter operations. Our approach (diagram in the middle) produces the, for this example, optimal solution: two loops. The first to calculate both sums, the second to calculate the maps.

>>>>

To demonstrate the effect of clustering and fusion, consider the following code snippet:
\begin{code}
 normalize :: Array Int -> Array Int
 normalize xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (> 0)    xs
        sum2 = fold   (+)  0   gts
        ys1  = map    (/ sum1) xs
        ys2  = map    (/ sum2) xs
    in (ys1, ys2)
\end{code}
The function @normalize@ computes two sums, one of all the elements of @xs@, the other of all the ones greater than zero. Since we need to fully evalute the sums to proceed with the maps, it is clear that we need at least two separate loops. These folds are examples of fusion-preventing dependencies, as fold cannot be fused with subsequent operations. Figure~\ref{f:normalize2-clusterings} shows the data-flow graph of @normalize@. In the leftmost diagram we can see the effect of applying stream fusion to the program: we end up with four loops (denoted by the dotted lines): only the filter operation is combined with the subsequent fold. The best existing ILP approach results in the right-most graph: it combines the @sum1@ fold and @sum2@ filter in one loop, but requires an extra loop for the fold operation which consumes the filter output, since it cannot fuse filter operations. Our approach (diagram in the middle) produces the, for this example, optimal solution: two loops. The first to calculate both sums, the second to calculate the maps.
